Modular Static Analysis: Orthogonal Design, Correct by Construction

### 2014-10-27 writeup/pldi15.tex
interpreter and exposing abstract domain content in.  now working on exposing flow sensitivities.

\newcommand{\ifz}[3]{\pmb{\operatorname{if0}} (#1)\{#2\}\{#3\}}
\newcommand{\lam}{\pmb{λ}}
\newcommand{\indor}{\;|\;}
\newcommand{\plus}{\pmb{+}}
\newcommand{\minus}{\pmb{-}}
\newcommand{\op}{\mathbin{op}}
\newcommand{\ap}{\;\;}
\newcommand{\steps}{\leadsto}
\newcommand{\ddo}{\underline{\operatorname{do}}}
\newcommand{\case}[1]{\text{\underline{case} $#1$ \underline{of}}}
\newcommand{\pushkon}{\operatorname{push-Kon}}
\newcommand{\popkon}{\operatorname{pop-Kon}}
\newcommand{\getenv}{\operatorname{get-Env}}
\newcommand{\putenv}{\operatorname{put-Env}}
\newcommand{\ret}{\underline{\operatorname{return}}}
\newcommand{\toclo}{\operatorname{to-clo}}
\newcommand{\mif}[3]{\text{\underline{if} $#1$ \underline{then} $#2$ \underline{else} $#3$}}
\newcommand{\stepf}{\operatorname{step}}
\newcommand{\atomf}{\operatorname{atom}}
\newcommand{\branchOnSet}{\operatorname{branch-on-set}}
\newcommand{\intI}{\operatorname{int-I}}
\newcommand{\ifzE}{\operatorname{if0-E}}
\newcommand{\cloI}{\operatorname{clo-I}}
\newcommand{\cloE}{\operatorname{clo-E}}
\newcommand{\st}{\;|\;}

@@

% Modular Analysis by Example {{{
\section{Modular Analysis by Example}

To demonstrate our framework we will grow a concrete semantics into an executable analysis in small steps.
Each step will expose some property of the analysis to be tuned independent of other properties.
We use an applied lambda calculus as an example language to demonstrate our approach.

% }}}

@@

\begin{figure}[t]
\caption{\lamif\ syntax.}
\label{lamif:syntax}
\input{snips/Lam-IF.tex}
\end{figure}

@@

\begin{figure}[t]
\caption{\lamif\ semantics.}
\label{lamif:semantics}
{\scriptsize
\input{snips/Lam-IF-Semantics-SS.tex}
\input{snips/Lam-IF-Semantics-Rules.tex}
}
\end{figure}

The analyses we will consider will be stated as the least fixed point of some abstract collecting semantics.
A collecting semantics for \lamif\ is defined as:
\begin{equation*}
  μ(ς). ς₀ ⊔ ς ⊔ \{ ς' \st ς \steps ς' \}
\end{equation*}
where $ς₀ := ⟨ e₀ , ⊥ , ∙ ⟩$ is the injection of the initial program $e₀$ into the state space $Σ$.

@@

% From Semantics to Interpreter {-{
\section{From Semantics to Interpreter}

On our way to an executable analysis for \lamif, we first evolve the small-step relation into an executable small-step interpreter.
We write this interpreter in monadic style for two reasons.
First, it is much easier to add new states and rules to a monadic interpreter.
Second, we will exploit the monadic abstraction to expose flow and path sensitivities in a later interpreter.

The monad for the interpreter will have two types effects: state and partiality.
Two operations carry state effects: get and put; and one operation carries a partiality effect: fail.

We construct a monad $ℳ $ which combines state effects for $Env$ and $Kon$ with partiality effects.
The type $ℳ $ is just a simple type which supports the monad, state, and partiality operations.
The interpreter for \lamif\ which uses $ℳ $ and monadic effects is given in figure~\ref{lamif:minterpreter}.

\begin{figure}[t]
\caption{\lamif\ monadic interpreter.}
\label{lamif:minterpreter}
{\scriptsize
\input{snips/Lam-IF-MInterpreter.tex}
}
\end{figure}

As before, we must construct a collecting semantics in order to define an executable analysis.
To do this, we must relate our step function back to some state machine step \emph{function}.
One we c % HERE

@@

% Exposing the Abstract Domain {-{
\section{Exposing the Abstract Domain}

@@

To expose the choice of abstract domain we introduce a type $AVal$ behind an abstract interface.
\begin{alignat*}{1}
    AVal &∈ Set                \\
   \intI &: ℤ → AVal           \\
   \ifzE &: AVal → 𝒫 (𝔹)       \\
   \cloI &: Clo → AVal         \\
   \cloE &: AVal → 𝒫 (Clo)     \\
  \delta &: Op → AVal → AVal → AVal
\end{alignat*}

To establish the correctness of our interpreter $\stepf$, we must prove that it is monotonic in $AVal$, $intI$, $if0E$, $cloI$ and $cloE$.
This can be done independent of a specific $AVal$.
Then we need only establish $Val \galois{α}{γ} AVal$ for some $AVal$ in order to justify the correctness of $\stepf$ instantiated with $AVal$.

Exposing the abstract domain forces us to reconsider our monadic effects.
For example, a branch on a possibly unknown integer value must return multiple possible machine states.
Therefore we must trade our partiality monad for a nonteterminism monad.
The monadic abstraction supports nondeterminism nicely in this setting.
The helper function $\branchOnSet$ will return multiple times for each value in the set.
This allows the interpreter to be written in a style that need only consider one state at a time.
The monadic interpreter whith the abstract domain exposd is given in figure~\ref{lamif:minterpreter:adom}.
(The implementation of $\stepf$ is unchanged.)

\begin{figure}[t]
\caption{\lamif\ monadic interpreter with abstract domain exposed.}
\label{lamif:minterpreter:adom}
{\scriptsize
\input{snips/Lam-IF-MInterpreter-ADom.tex}
}
\end{figure}

>>>>

\newcommand{\ifz}[3]{\pmb{\operatorname{if0}} (#1)\{#2\}\{#3\}}
\newcommand{\lam}{\pmb{λ}}
\newcommand{\indor}{\;|\;}
\newcommand{\plus}{\pmb{+}}
\newcommand{\minus}{\pmb{-}}
\newcommand{\op}{\mathbin{op}}
\newcommand{\ap}{\;\;}
\newcommand{\steps}{\leadsto}
\newcommand{\ddo}{\underline{\operatorname{do}}}
\newcommand{\case}[1]{\text{\underline{case} $#1$ \underline{of}}}
\newcommand{\getkon}{\operatorname{get-Kon}}
\newcommand{\putkon}{\operatorname{put-Kon}}
\newcommand{\pushkon}{\operatorname{push-Kon}}
\newcommand{\popkon}{\operatorname{pop-Kon}}
\newcommand{\getenv}{\operatorname{get-Env}}
\newcommand{\putenv}{\operatorname{put-Env}}
\newcommand{\ret}{\operatorname{return}}
\newcommand{\toclo}{\operatorname{to-clo}}
\newcommand{\mif}[3]{\text{\underline{if} $#1$ \underline{then} $#2$ \underline{else} $#3$}}
\newcommand{\stepf}{\operatorname{step}}
\newcommand{\atomf}{\operatorname{atom}}
\newcommand{\branchOnSet}{\operatorname{branch-on-set}}
\newcommand{\intI}{\operatorname{int-I}}
\newcommand{\ifzE}{\operatorname{if0-E}}
\newcommand{\cloI}{\operatorname{clo-I}}
\newcommand{\cloE}{\operatorname{clo-E}}
\newcommand{\aintI}{\widehat{\operatorname{int-I}}}
\newcommand{\aifzE}{\widehat{\operatorname{if0-E}}}
\newcommand{\acloI}{\widehat{\operatorname{clo-I}}}
\newcommand{\acloE}{\widehat{\operatorname{clo-E}}}
\newcommand{\adelta}{\widehat{\delta}}
\newcommand{\st}{\;|\;}
\newcommand{\coerce}[1]{\operatorname{↓_{#1}}}
\newcommand{\failonbot}{\operatorname{↑_⊥}}
\newcommand{\liftnondet}{\operatorname{↑_𝒫 }}
\newcommand{\AVal}{\widehat{Val}}
\newcommand{\AEnv}{\widehat{Env}}
\newcommand{\mplus}{\mathbin{⟨+⟩}}

@@

% Modular Analysis by Example {-{
\section{Modular Analysis by Example}

To demonstrate our framework we will grow a concrete semantics into an executable analysis in small steps.
% Should I resist the pun? :P -DD
Each step will expose some property of the analysis to be tuned independent of other properties.
We use an applied lambda calculus as an example language to demonstrate our approach.

@@

\begin{figure}[t]
\caption{\lamif\ syntax.}
\label{lamif:syntax}
\input{snips/00_Semantics/00_Syntax.tex}
\end{figure}

@@

\begin{figure}[t]
\caption{\lamif\ semantics.}
\label{lamif:semantics}
\input{snips/00_Semantics/01_Semantics.tex}
\end{figure}

Analyses in our framework are least fixed point computations over a collecting semantics.
A collecting semantics for \lamif\ is defined as:
\begin{equation*}
  μ(ς). ς₀ ⊔ ς ⊔ \{ ς' \st ς \steps ς' \}
\end{equation*}
where $ς₀ := ⟨ e₀ , ⊥ , ∙ ⟩$ is the injection of the initial program $e₀$ into the state space $Σ$.

@@

% From Semantics to Interpreter {-{
\subsection{From Semantics to Interpreter}
\label{interpreter}

On our way to an executable analysis for \lamif, we first evolve the small-step relation into an executable small-step interpreter.
We write this interpreter in monadic style for two reasons.
First, it is much easier to add new states and rules to a monadic interpreter.
Second, we will exploit the monadic abstraction to expose flow and path sensitivities in a later section.

The interpreter will use two types monadic effects: state and partiality.
The operations for state effects are get and put, and the operation for the partiality effect is $⊥$.
For reference: monadic, state and partiality operations have the following types:

\input{snips/01_Interpreter/00_MonadOps.tex}
%\input{snips/01_Interpreter/01_StateOps.tex}
%\input{snips/01_Interpreter/02_PartialityOps.tex}

We construct a monad $ℳ $ for our interpreter which combines state effects for $Env$ and $Kon$ with a partiality effect.
The type $ℳ $ is just a simple type:
\input{snips/01_Interpreter/03_MonadType.tex}

We factor the definition of the interpreter into two parts: one for atomic and one for compound expressions.
Atomic expressions return to the top frame of the stack:
\input{snips/01_Interpreter/04_Atomic.tex}

Our interpreter uses helper functions $𝒜 ⟦\_⟧$, $\pushkon$, $\popkon$, $\delta$, $\coerce{Clo}$, and $\failonbot$.
The function $𝒜 ⟦\_⟧$ is the denotation function which uses the monadic state effect for environments:
\input{snips/01_Interpreter/05_AtomDen.tex}

Functions $\pushkon$ and $\popkon$ interact with effects for both continuation state and partiality:
\input{snips/01_Interpreter/06_PushPopKon.tex}

The function $\delta$ gives the expected denotation to operators:
\input{snips/01_Interpreter/07_Delta.tex}

The function $\coerce{Clo}$ is a function from types which include closure values to partial closures.
\input{snips/01_Interpreter/08_CoerceClo.tex}

The function $\failonbot$ lifts partial values to monadic values, mapping $⊥$ to $⊥$.
\input{snips/01_Interpreter/09_FailOnBot.tex}

Our small-step interpreter is defined by dispatching to $\atomf$ or pushing continuation frames for compound expressions.
\input{snips/01_Interpreter/10_Exp.tex}

\paragraph{Correctness and Execution}

As before, we must construct a collecting semantics in order to define an executable analysis.
To do this, we must relate our small-step interpreter to a state machine transition function.
That is, we must relate our interpreter of type $Exp → ℳ (Exp)$ to a state transition of type $Σ → Σ$ for \emph{some} $Σ$.

For this interpreter the state space $Σ$ is defined $Σ := {(Exp × Env × Kon)}_⊥$.
To establish the correctness of $Σ$ we construct an isomorphism between $Exp → ℳ (Exp)$ and $Σ → Σ$.
We use the standard isomorphisim between Kleisli morphisms $α → ℳ (β)$ and $ℳ (α) → ℳ (β)$ where $ℳ $ is the partiality monad ${(\_)}_⊥$.
To execute the analysis we transport the interpreter through the isomorphism and find the least fixed point of the collecting semantics.

@@

% Exposing the Abstract Domain {-{
\subsection{Exposing the Abstract Domain}

@@

To expose the choice of abstract domain we introduce a type $\AVal$ behind an abstract interface:
\input{snips/02_AbstractDomain/00_AVal.tex}

This abstraction forces us to use an abstract environment, mapping variables to abstract values:
\input{snips/02_AbstractDomain/01_AEnv.tex}
and re-implement $𝒜 ⟦\_⟧$:
\input{snips/02_AbstractDomain/02_AtomDen.tex}

Before implementing $\atomf$ we must reconsider our monadic effects.
Because we have abstracted values, our interpreter must consider \emph{sets} of possible states resulting from conditionals and function calls.
We introduce this nondeterminism as an effect in the monad, taking the place of partiality (which was just nondeterminism with zero or one branches).
Nondeterminism has two operators, $⊥$ as we've seen before, and $\mplus$ which branches to both left and right monadic computation.
The effects for our monadic type are now:
\input{snips/02_AbstractDomain/03_Effects.tex}
and to support these effects we redefine $ℳ $ to be:
\input{snips/02_AbstractDomain/04_MonadType.tex}

Our definition for $\atomf$ now uses nondeterminism effects:
\input{snips/02_AbstractDomain/05_Atomic.tex}

Helper function $\liftnondet$ is the analog of $\failonbot$ for lifting sets into monads with nondeterminism:
\input{snips/02_AbstractDomain/06_LiftNonDet.tex}

\paragraph{Correctness and Execution}

To construct a collecting semantics we again construct an isomorphism between $Exp → ℳ (Exp)$ and $Σ$ where $Σ := 𝒫 (Exp × \AEnv × Kon)$.
This isomorphism is again the standard one between Kleisli morphisms $α → ℳ (β)$ and $ℳ (α) → ℳ (β)$ where $ℳ $ is the powerset monad $𝒫 $.

Now that we have a parameter to the analysis, $\AVal$, we can discuss what an analysis is and what it means for it to be correct.
An analysis is some instantiation of $\AVal$ to $\AVal₁$ where $Val\galois{α}{γ}\AVal₁$.
Furthermore, each operation in the $\AVal$ interface must be an abstraction of concrete versions, namely $\intI ⊑ γ (\aintI₁)$, $\ifzE ⊑ γ (\aifzE₁)$, etc.
This local argument about $\AVal₁$, coupled with the monotonicity of $\stepf$ and the isomorphism between $Exp → ℳ (Exp)$ and $Σ → Σ$, 
  gives us an end to end correctness argument for our executable abstract interpreter.
At this point our analysis is still not yet decidable; the abstract state space is still infinite.

We will expose two more feature, flow sensitivity and context sensitivity, before we achieve a decidable analysis.
However, the \emph{correctness} of our analysis will be carried through every example.

% Exposing Flow Sensitivity {-{
\subsection{Exposing Flow Sensitivity}

Novel in this work is the ability to expose flow sensitivity as an orthogonal parameter to a static analysis.

% Exposing Context Sensitivity {-{
\subsection{Exposing Context Sensitivity}

% Intentional Optimizations {-{
\subsection{Intentional Optimizations}

\subsubsection{Abstract Garbage Collection}

\subsubsection{Flat Closures}

% Summary {-{
\section{Summary}

Our final interpreter enjoys the following properties:
\begin{itemize}
\item Abstract Domain can be altered independent of other features.
\item Flow Sensitivity can be altered independent of other features.
\item Context Sensitivity (and its granularity) can be altered indepndend of other features.
\item Intensional optimizations can be implemented once and integrate seamlessly with any of the above choices.
\item New state space components and their widening properties can be added seamlessly using compositional building blocks.
\item Interpreters are automatically related to a collecting semantics, enabling a straight forward least-fixed-point implementation strategy.
\item All analyses are correct by construction with no additional proof burdon beyond interpreter monotonicity and correctness of orthoganal parameters piecewise.
\end{itemize}
